graphsim: An R package for simulating gene expression data from graph structures of biological pathways

### 2020-05-08 paper/paper.tex
correct figure widths (fig2-4, pane E)

\documentclass[article, shortnames]{jss}

%% -- LaTeX packages and custom commands --------------

%% recommended packages
\usepackage{thumbpdf,lmodern}

%% another package (only for this demo article)
\usepackage{framed}
\usepackage{subcaption} %continuefloat
\usepackage{xfrac}
\usepackage{mathtools}
\usepackage{natbib}

\usepackage[acronym, xindy, toc, acronym, style=super,  nogroupskip, nonumberlist, nopostdot]{glossaries}
\setlength{\glsdescwidth}{0.75\textwidth}
\include{glossary}
\makeglossaries

\usepackage{placeins}

%\usepackage[colorlinks,
%           linkcolor={black},
%           citecolor={black},
%           urlcolor={blue!80!white}]{hyperref}
           %allbordercolors={0 0 0},
           %pdfborderstyle={/S/U/W 1}

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

%% - Article metainformation (author, title, ...) ----------

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{S. Thomas Kelly\\University of Otago
   \And Michael A. Black\\University of Otago}
\Plainauthor{Thomas Kelly, Michael Black}

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{\pkg{graphsim}: An \proglang{R} package for simulating gene expression data from graph structures of biological pathways}
\Plaintitle{graphsim: An R package for simulating gene expression data from graph structures of biological pathways}
\Shorttitle{Simulating gene expression from pathways in R}

%% - \Abstract{} almost as usual
\Abstract{
Transcriptomic analysis is used to capture the molecular state of a cell or sample in many biological and medical applications. In addition to identifying alterations in activity at the level of individual genes, understanding changes in the gene networks that regulate fundamental biological mechanisms is also an important objective of molecular analysis. As a result, databases that describe biological pathways are increasingly relied on to assist with the interpretation of results from large-scale genomics studies. Incorporating information from biological pathways and gene regulatory networks into a genomic data analysis is a popular strategy, and there are many methods that provide this functionality for gene expression data. When developing or comparing such methods, it is important to gain an accurate assessment of their performance, with simulation-based validation studies a popular choice.  This necessitates the use of simulated data that correctly accounts for pathway relationships and correlations. Here we present a versatile statistical framework to simulate correlated gene expression data from biological pathways, by sampling from a multivariate normal distribution derived from a graph structure. This procedure has been released as the \pkg{graphsim} \proglang{R} package (\url{https://github.com/TomKellyGenetics/graphsim}) and is compatible with any graph structure that can be described using the \pkg{igraph} package.
}

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{gene-expression, simulation, genomics, pathway, network, \proglang{R}}
\Plainkeywords{gene-expression, simulation, genomics, pathway, network, R}

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  S. Thomas Kelly\\
  Division of Genomic Medicine
  RIKEN Center for Integrative Medical Sciences\\
  Yokohama, Japan \\
  E-mail: \email{tom.kelly@riken.jp}\\
  \emph{and}\\
  Department of Biochemistry\\
  University of Otago\\
  Dunedin, New Zealand\\
  %E-mail: \email{tom.kelly@postgrad.otago.ac.nz}%\\
  %URL: \url{https://eeecon.uibk.ac.at/~zeileis/}

  Michael Black\\
  Department of Biochemistry\\
  University of Otago\\
  PO Box 56\\
  Dunedin, New Zealand\\
  E-mail: \email{mik.black@otago.ac.nz}%\\
  %URL: \url{https://eeecon.uibk.ac.at/~zeileis/}
}

\begin{document}

%% - Introduction ---------------------

%% - In principle "as usual".
%% - But should typically have some discussion of both _software_ and _methods_.
%% - Use \proglang{}, \pkg{}, and \code{} markup throughout the manuscript.
%% - If such markup is in (sub)section titles, a plain text version has to be
%%   added as well.
%% - All software mentioned should be properly \cite-d.
%% - All abbreviations should be introduced.
%% - Unless the expansions of abbreviations are proper names (like "Journal
%%   of Statistical Software" above) they should be in sentence case (like
%%   "generalized linear models" below).

\section[Introduction: inference and modelling of biological networks]{Introduction: inference and modelling of biological networks} \label{sec:intro}

\shortcites{Perou2000, Arner2015, Reactome, Komatsu2013, Yamaguchi2007, Shimamura2009, Hirose2008, Li2015, Genz2016}

Network analysis of molecular biological pathways has the potential to lead to new insights into biology and medical genetics \citep{Barabasi2004, Hu2016}. Since gene expression profiles capture a consistent signature of the regulatory state of a cell \citep{Perou2000, Ozsolak2011, Svensson2018}, they can be used to analyse complex molecular states with genome-scale data. However, biological pathways are often analysed in a reductionist paradigm as amorphous sets of genes involved in particular functions, despite the fact that the relationships defined by pathway structure could further inform gene expression analyses. In many cases, the pathway relationships are well-defined, experimentally-validated, and are available in public databases \citep{Reactome}. 
%While pathway analysis over-representation analysis using the hypergeometric test is an effective approach to identify enriched functional groups \citep{Mooney2015, Werner2008}, it is limited by the accuracy and coverage of the biological databases used. Furthermore, pathway enrichment does not utilise the pathway structure information that has been gathered for many pathways and is supported by databases such as Reactome \citep{Reactome}. 
As a result, network analysis techniques could play an important role in furthering our understanding of biological pathways and aiding in the interpretation of genomics studies. %Furthermore with modelling and simulation procedures to evaluate them are needed.

Gene networks provide insights into how cells are regulated, by mapping regulatory interactions between target genes and transcription factors, enhancers, and sites of epigenetic marks or chromatin structures \citep{Barabasi2004, Yamaguchi2007}. Inference of these regulatory interactions for genomics investigations has the potential to radically expand the range of candidate biological pathways to be further explored, or to improve the accuracy of bioinformatics and functional genomic analysis. %which relies on these pathways being curated.
A number of methods have already been developed to utilise timecourse gene expression data \citep{Arner2015, Yamaguchi2007} using gene regulatory modules in state-space models and recursive vector autoregressive models \citep{Hirose2008, Shimamura2009}. Various approaches to gene regulation and networks at the genome-wide scale have lead to novel biological insights \citep{Arner2015, Komatsu2013}. However, inference of regulatory networks has thus far relied on experimental validation or resampling-based approaches to estimate the likelihood of specific network modules being predicted \citep{Markowetz2007,Hawe2019}.

There is a need, therefore, for a systematic framework for statistical modelling and simulation of gene expression data derived from hypothetical, inferred or known gene networks. Here we present an \proglang{R} package to achieve this, where samples from a multivariate normal distribution are used to generate normally-distributed log-expression data, with correlations between genes derived from the structure of the underlying pathway or gene regulatory network. This methodology enables simulation of expression profiles that approximate the log-transformed and normalised data from microarray and bulk or single-cell RNA-Seq experiments. This procedure has been released as the \pkg{graphsim} \proglang{R} package to enable the generation of simulated gene expression datasets containing pathway relationships from a known underlying network. These simulated datasets can be used to evaluate various bioinformatics methodologies, including statistical and network inference procedures.

%This procedure enables modelling of genes within a biological pathway and the investigation of the impact of pathway structure on identification of biomarkers from gene expression analysis techniques. It was developed for biological pathway members with correlated gene expression (higher than the background of genes in other pathways) but it may also be applicable to modelling protein levels (e.g, in a kinase regulation cascade) or substrates and products (e.g., in a metabolic pathway). The simulation procedure was designed to use graph structures (such as those shown in Figure~\ref{fig:simple_graph}) to inform development of a ``Sigma'' variance-covariance matrix ($\Sigma$) for sampling from a multivariate normal distribution. % (using the \pkg{mvtnorm} \proglang{R} package).

\iffalse
\begin{leftbar}
The introduction is in principle ``as usual''. However, it should usually embed
both the implemented \emph{methods} and the \emph{software} into the respective
relevant literature. For the latter both competing and complementary software
should be discussed (within the same software environment and beyond), bringing
out relative (dis)advantages. All software mentioned should be properly
\verb|\cite{}|d. (See also Appendix~\ref{app:bibtex} for more details on
\textsc{Bib}{\TeX}.)

For writing about software JSS requires authors to use the markup
\verb|\proglang{}| (programming languages and large programmable systems),
\verb|\pkg{}| (software packages), \verb|\code{}| (functions, commands,
arguments, etc.). If there is such markup in (sub)section titles (as above), a
plain text version has to be provided in the {\LaTeX} command as well. Below we
also illustrate how abbreviations should be introduced and citation commands can
be employed. See the {\LaTeX} code for more details.
\end{leftbar}

Modeling count variables is a common task in economics and the social sciences.
The classical Poisson regression model for count data is often of limited use in
these disciplines because empirical count data sets typically exhibit
overdispersion and/or an excess number of zeros. The former issue can be
addressed by extending  the plain Poisson regression model in various
directions: e.g., using sandwich covariances or estimating an additional
dispersion parameter (in a so-called quasi-Poisson model). Another more formal
way is to use a negative binomial (NB) regression. All of these models belong to
the family of generalized linear models (GLMs). However, although these models
typically can capture overdispersion rather well, they are in many applications
not sufficient for  modeling excess zeros. Since \citep{Mullahy:1986} there is
increased interest in zero-augmented models that address this issue by a second
model component capturing zero counts. An overview of count data models in
econometrics, including  hurdle and zero-inflated models, is provided in
\citep{Cameron+Trivedi:2013}.

In \proglang{R} \citep{R}, GLMs are provided by the model fitting functions
\fct{glm} in the \pkg{stats} package and \fct{glm.nb} in the \pkg{MASS} package
\citep[][Chapter~7.4]{Venables+Ripley:2002} along with associated methods for
diagnostics and inference. The manuscript that this document is based on
\citep{Zeileis+Kleiber+Jackman:2008} then introduced hurdle and zero-inflated
count models in the functions \fct{hurdle} and \fct{zeroinfl} in the \pkg{pscl}
package \citep{Jackman:2015}. Of course, much more software could be discussed
here, including (but not limited to) generalized additive models for count data
as available in the \proglang{R} packages \pkg{mgcv} \citep{Wood:2006}, 
\pkg{gamlss} \citep{Stasinopoulos+Rigby:2007}, or \pkg{VGAM} \citep{Yee:2009}.
\fi

%% - Manuscript ---------------------

%% - In principle "as usual" again.
%% - When using equations (e.g., {equation}, {eqnarray}, {align}, etc.
%%   avoid empty lines before and after the equation (which would signal a new
%%   paragraph.
%% - When describing longer chunks of code that are _not_ meant for execution
%%   (e.g., a function synopsis or list of arguments), the environment {Code}
%%   is recommended. Alternatively, a plain {verbatim} can also be used.
%%   (For executed code see the next section.)

\section{Methodology and software} \label{sec:methods}

\begin{figure*}[!htb]
%\begin{mdframed}
%  \resizebox{\textwidth}{!}{
         \begin{center}
%
        \subcaptionbox{Activating pathway structure  \label{fig:simple_graph:first}}{%
            \fbox{
            \includegraphics[width=0.45\textwidth]{{simple_graph.png}}
            }
        }%
        \subcaptionbox{Pathway structure with inhibitions \label{fig:simple_graph:second}}{%
            \fbox{
           \includegraphics[width=0.45\textwidth]{{simple_graph_inhibiting.png}}
           }
        }%
%
    \end{center}
   \caption[Simulated graph structures]{\small \textbf{\textbf{Simulated graph structures.}} A constructed graph structure used as an example to demonstrate the simulation procedure in Figures~\ref{fig:simulation_activating} and~\ref{fig:simulation_inhibiting}. Activating links are denoted by black arrows and inhibiting links by red edges. Inhibiting edges have been highlighted in red.}
%}
\label{fig:simple_graph}
%\end{mdframed}
\end{figure*}

Here we present a procedure to simulate gene expression data with correlation structure derived from a known graph structure.  This procedure assumes that transcriptomic data have been generated and follow a log-normal distribution (i.e., $log(X_{ij}) \sim MVN({\bf\mu}, \Sigma)$, where ${\bf\mu}$ and $\Sigma$ are the mean vector  and variance-covariance matrix respectively, for gene expression data derived from a biological pathway) after appropriate normalisation \citep{Law2014, Li2015}. Log-normality of gene expression matches the assumptions of the popular \pkg{limma} package, which is often used for the analysis of intensity-based data from gene expression microarray studies and count-based data from RNA-Seq experiments. This approach has also been applied for modelling UMI-based count data from single-cell RNA-Seq experiments in the \pkg{DESCEND} package \citep{Wang2018}.
%rather than correlated blocks for $\Sigma$ in \pkg{mvtnorm} \citet{mvtnorm}
%This procedure produces a ``Sigma'' variance-covariance matrix ($\Sigma$) for sampling from a multivariate normal distribution (using the \pkg{mvtnorm} \proglang{R} package).

In order to simulate transcriptomic data, a pathway is first constructed as a graph structure, using the \pkg{igraph} \proglang{R} package \citep{igraph}, with the status of the edge relationships defined (i.e, whether they activate or inhibit downstream pathway members). 
%
\textcolor{black}{This procedure uses} a graph structure such as that presented in Figure~\ref{fig:simple_graph:first}. The graph can be defined by an adjacency matrix, \textbf{$A$} (with elements \textcolor{black}{$A_{ij}$}), where  \textcolor{black}{
%
\[
A_{ij} = 
\begin{dcases*}
   1                         & if genes $i$ and $j$ are adjacent \\
   0                         & otherwise
\end{dcases*}
\]}
%
\noindent
A matrix, \textbf{$R$}, with elements \textcolor{black}{$R_{ij}$}, is calculated based on distance (i.e., number of edges contained in the shortest path) between nodes, such that closer nodes are given more weight than more distant nodes, to define inter-node relationships. A geometrically-decreasing (relative) distance weighting is used to achieve this: \textcolor{black}{
%
\[
R_{ij} = 
\begin{dcases*}
   1                            & if genes $i$ and $j$ are adjacent \\
   (\sfrac{1}{2})^{d_{ij}}      & if a path can be found  between genes $i$ and $j$ \\
   0                            & if no path exists between genes $i$ and $j$ 
\end{dcases*}
\]}
%
\noindent
where $d_{ij}$ is the length of the shortest path (i.e., minimum number of edges traversed) between genes (nodes) $i$ and $j$ in graph $G$. Each more distant node is thus related by $\sfrac{1}{2}$ compared to the next nearest, as shown in Figure~\ref{fig:simulation_activating:second}.
An arithmetically-decreasing (absolute) distance weighting is also supported in the \pkg{graphsim} \proglang{R} package which implements this procedure: \textcolor{black}{
%
\[
R_{ij} = 
\begin{dcases*}
   1                            & if genes $i$ and $j$ are adjacent \\
   1-\frac{d_{ij}}{diam(G)}     & if a path can be found  between genes $i$ and $j$ \\
   0                            & if no path exists between genes $i$ and $j$ 
\end{dcases*}
\]
%
}

Assuming a unit variance for each gene, these values can be used to derive a $\Sigma$ matrix:
%
\[
\Sigma_{ij} = 
\begin{dcases*}
   1                            & if $i=j$ \\
   \rho R_{ij}  & otherwise
\end{dcases*}
\]
%
\noindent
where $\rho$ is the correlation between adjacent nodes. 
%The $\Sigma$ matrix (with a diagonal of $1$) would thus have a standard deviation of 1 for each node. 
Thus covariances between adjacent nodes are assigned by a correlation parameter ($\rho$) and the remaining off-diagonal values in the matrix are based on scaling these correlations by the geometrically weighted relationship matrix (or the nearest positive definite matrix for $\Sigma$ with negative correlations).\\

Computing the nearest positive definite matrix is necessary to ensure that the variance-covariance matrix could be inverted when used as a parameter in multivariate normal simulations, particularly when negative correlations are included for inhibitions (as shown below). Matrices that could not be inverted occurred rarely with biologically plausible graph structures but this approach allows for the computation of a plausible correlation matrix when the graph structure given is incomplete or contains loops. When required, the nearest positive definite matrix is computed using the \texttt{nearPD} function of the \pkg{Matrix} \proglang{R} package \citep{Matrix} to perform Higham's algorithm \citep{Higham2002} on variance-covariance matrices. The \pkg{graphsim} package gives a warning when this occurs.

\section{Illustrations} \label{sec:illustrations}
\subsection{Generating a Graph Structure} \label{sec:plot_graph}

The graph structure in Figure~\ref{fig:simple_graph:first} was used to simulate correlated gene expression data by sampling from a multivariate normal distribution using the \pkg{mvtnorm} \proglang{R} package \citep{Genz2009, mvtnorm}.
%This simulation procedure will be demonstrated with the relatively simple constructed graph structure shown in Figure~\ref{fig:simple_graph}. 
The graph structure visualisation in Figure~\ref{fig:simple_graph} was specifically developed for (directed) iGraph objects in \proglang{R} and is available in the \pkg{plot.igraph} and \pkg{igraph.extensions} packages. The \texttt{plot\_directed} function enables customisation of plot parameters for each node or edge, and mixed (directed) edge types for indicating activation or inhibition. These inhibition links (which occur frequently in biological pathways) are demonstrated in Figure~\ref{fig:simple_graph:second}.

@@

\begin{CodeChunk}
\begin{CodeInput}
#install packages required (once per machine)
install.packages("igraph")
install.packages("devtools")
library("devtools")
devtools::install\_github("TomKellyGenetics/graphsim")
#install custom plotting package
devtools::install\_github("TomKellyGenetics/plot.igraph")
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
#load required packages (once per R instance)
library("igraph")
library("graphsim")
library("plot.igraph")
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
#generate graph structure
graph\_edges <- rbind(c("A", "C"), c("B", "C"), c("C", "D"), c("D", "E"), 
    c("D", "F"), c("F", "G"), c("F", "I"), c("H", "I"))
graph <- graph.edgelist(graph\_edges, directed = T)

#plot graph structure (Figure 1)
plot\_directed(graph, state = "activating", layout = layout.kamada.kawai, 
    cex.node=3, cex.arrow=5, arrow\_clip = 0.2)
\end{CodeInput}
\end{CodeChunk}

>>>>

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={graphsim: An R package for simulating gene expression data from graph structures of biological pathways},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{caption}

\title{graphsim: An R package for simulating gene expression data from graph
structures of biological pathways}
\author{}
\date{\vspace{-2.5em}30 April 2020}

\begin{document}
\maketitle

\hypertarget{summary}{%
\subsubsection{Summary}\label{summary}}

Transcriptomic analysis is used to capture the molecular state of a cell
or sample in many biological and medical applications. In addition to
identifying alterations in activity at the level of individual genes,
understanding changes in the gene networks that regulate fundamental
biological mechanisms is also an important objective of molecular
analysis. As a result, databases that describe biological pathways are
increasingly relied on to assist with the interpretation of results from
large-scale genomics studies. Incorporating information from biological
pathways and gene regulatory networks into a genomic data analysis is a
popular strategy, and there are many methods that provide this
functionality for gene expression data. When developing or comparing
such methods, it is important to gain an accurate assessment of their
performance, with simulation-based validation studies a popular choice.
This necessitates the use of simulated data that correctly accounts for
pathway relationships and correlations. Here we present a versatile
statistical framework to simulate correlated gene expression data from
biological pathways, by sampling from a multivariate normal distribution
derived from a graph structure. This procedure has been released as the
\texttt{graphsim} R package on CRAN and GitHub
(\url{https://github.com/TomKellyGenetics/graphsim}) and is compatible
with any graph structure that can be described using the \texttt{igraph}
package. This package allows the simulation of biological pathways from
a graph structure based on a statistical model of gene expression.

\hypertarget{sec:intro}{%
\section{Introduction: inference and modelling of biological
networks}\label{sec:intro}}

Network analysis of molecular biological pathways has the potential to
lead to new insights into biology and medical genetics (Barabási and
Oltvai 2004; Hu, Thomas, and Brunak 2016). Since gene expression
profiles capture a consistent signature of the regulatory state of a
cell (Perou et al. 2000; Ozsolak and Milos 2011; Svensson, Vento-Tormo,
and Teichmann 2018), they can be used to analyse complex molecular
states with genome-scale data. However, biological pathways are often
analysed in a reductionist paradigm as amorphous sets of genes involved
in particular functions, despite the fact that the relationships defined
by pathway structure could further inform gene expression analyses. In
many cases, the pathway relationships are well-defined,
experimentally-validated, and are available in public databases (Croft
et al. 2014). As a result, network analysis techniques could play an
important role in furthering our understanding of biological pathways
and aiding in the interpretation of genomics studies.

Gene networks provide insights into how cells are regulated, by mapping
regulatory interactions between target genes and transcription factors,
enhancers, and sites of epigenetic marks or chromatin structures
(Barabási and Oltvai 2004; Yamaguchi et al. 2007). Inference of these
regulatory interactions for genomics investigations has the potential to
radically expand the range of candidate biological pathways to be
further explored, or to improve the accuracy of bioinformatics and
functional genomic analysis. A number of methods have already been
developed to utilise timecourse gene expression data (Arner et al. 2015;
Yamaguchi et al. 2007) using gene regulatory modules in state-space
models and recursive vector autoregressive models (Hirose et al. 2008;
Shimamura et al. 2009). Various approaches to gene regulation and
networks at the genome-wide scale have led to novel biological insights
(Arner et al. 2015; Komatsu et al. 2013). However, inference of
regulatory networks has thus far relied on experimental validation or
resampling-based approaches to estimate the likelihood of specific
network modules being predicted (Markowetz and Spang 2007; Hawe, Theis,
and Heinig 2019).

There is a need, therefore, for a systematic framework for statistical
modelling and simulation of gene expression data derived from
hypothetical, inferred or known gene networks. Here we present a package
to achieve this, where samples from a multivariate normal distribution
are used to generate normally-distributed log-expression data, with
correlations between genes derived from the structure of the underlying
pathway or gene regulatory network. This methodology enables simulation
of expression profiles that approximate the log-transformed and
normalised data from microarray and bulk or single-cell RNA-Seq
experiments. This procedure has been released as the package to enable
the generation of simulated gene expression datasets containing pathway
relationships from a known underlying network. These simulated datasets
can be used to evaluate various bioinformatics methodologies, including
statistical and network inference procedures.

\hypertarget{sec:methods}{%
\section{Methodology and software}\label{sec:methods}}

Here we present a procedure to simulate gene expression data with
correlation structure derived from a known graph structure. This
procedure assumes that transcriptomic data have been generated and
follow a log-normal distribution (i.e.,
\(log(X_{ij}) \sim MVN({\bf\mu}, \Sigma)\), where \({\bf\mu}\) and
\(\Sigma\) are the mean vector and variance-covariance matrix
respectively, for gene expression data derived from a biological
pathway) after appropriate normalisation (Law et al. 2014; Li et al.
2015). Log-normality of gene expression matches the assumptions of the
popular \texttt{limma} package (Ritchie et al. 2015), which is often
used for the analysis of intensity-based data from gene expression
microarray studies and count-based data from RNA-Seq experiments. This
approach has also been applied for modelling UMI-based count data from
single-cell RNA-Seq experiments in the package (Wang et al. 2018).

In order to simulate transcriptomic data, a pathway is first constructed
as a graph structure, using the \texttt{igraph} R package (Csardi and
Nepusz 2006), with the status of the edge relationships defined (i.e,
whether they activate or inhibit downstream pathway members). {This
procedure uses} a graph structure such as that presented in
Figure~\protect\hyperlink{fig:simple_graph:first}{1a}. The graph can be
defined by an adjacency matrix, \textbf{\(A\)} (with elements
\(A_{ij}\)), where \[A_{ij} = 
\begin{cases}
   1                         & \mbox{if genes } i \mbox{ and } j \mbox{ are adjacent} \\
   0                         & \mbox{otherwise}
\end{cases}\]

A matrix, \textbf{\(R\)}, with elements {\(R_{ij}\)}, is calculated
based on distance (i.e., number of edges contained in the shortest path)
between nodes, such that closer nodes are given more weight than more
distant nodes, to define inter-node relationships. A
geometrically-decreasing (relative) distance weighting is used to
achieve this:

{\[R_{ij} = 
\begin{cases}
   1                            & \mbox{if genes } i \mbox{ and } j \mbox{ are adjacent} \\
   (\frac{1}{2})^{d_{ij}}      & \mbox{if a path can be found between genes } i \mbox{ and } j \\
   0                            & \mbox{if no path exists between genes } i \mbox{ and } j 
\end{cases}\]} where \(d_{ij}\) is the length of the shortest path
(i.e., minimum number of edges traversed) between genes (nodes) \(i\)
and \(j\) in graph \(G\). Each more distant node is thus related by
\(\frac{1}{2}\) compared to the next nearest, as shown in
Figure~\protect\hyperlink{fig:simulation_activating:second}{2b}. An
arithmetically-decreasing (absolute) distance weighting is also
supported in the \texttt{graphsim} R package which implements this
procedure: {\[R_{ij} = 
\begin{cases}
   1                            & \mbox{if genes } i \mbox{ and } j \mbox{ are adjacent} \\
   1-\frac{d_{ij}}{diam(G)}     & \mbox{if a path can be found between genes } i \mbox{ and } j \\
   0                            & \mbox{if no path exists between genes } i \mbox{ and } j 
\end{cases}\]}

Assuming a unit variance for each gene, these values can be used to
derive a \(\Sigma\) matrix: \[\Sigma_{ij} = 
\begin{cases}
   1                            & \mbox{if } i=j \\
   \rho R_{ij}  & \mbox{otherwise}
\end{cases}\] where \(\rho\) is the correlation between adjacent nodes.
Thus covariances between adjacent nodes are assigned by a correlation
parameter (\(\rho\)) and the remaining off-diagonal values in the matrix
are based on scaling these correlations by the geometrically weighted
relationship matrix (or the nearest positive definite matrix for
\(\Sigma\) with negative correlations).\\
Computing the nearest positive definite matrix is necessary to ensure
that the variance-covariance matrix could be inverted when used as a
parameter in multivariate normal simulations, particularly when negative
correlations are included for inhibitions (as shown below). Matrices
that could not be inverted occurred rarely with biologically plausible
graph structures but this approach allows for the computation of a
plausible correlation matrix when the given graph structure is
incomplete or contains loops. When required, the nearest positive
definite matrix is computed using the \texttt{nearPD} function of the
\texttt{Matrix} R package (Bates and Maechler 2016) to perform Higham's
algorithm (Higham 2002) on variance-covariance matrices. The
\texttt{graphsim} package gives a warning when this occurs.

\hypertarget{sec:illustrations}{%
\section{Illustrations}\label{sec:illustrations}}

\hypertarget{sec:plot_graph}{%
\subsection{Generating a Graph Structure}\label{sec:plot_graph}}

The graph structure in
Figure~\protect\hyperlink{fig:simple_graph:first}{1a} was used to
simulate correlated gene expression data by sampling from a multivariate
normal distribution using the R package (Genz and Bretz 2009; Genz et
al. 2016). The graph structure visualisation in
Figure~\protect\hyperlink{fig:simple_graph}{1} was specifically
developed for (directed) \texttt{igraph} objects in and is available in
the and packages. The \textbackslash{}texttt\{plot\_directed\} function
enables customisation of plot parameters for each node or edge, and
mixed (directed) edge types for indicating activation or inhibition.
These inhibition links (which occur frequently in biological pathways)
are demonstrated in
Figure~\protect\hyperlink{fig:simple_graph:second}{1b}.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{#install packages required (once per machine)}

\KeywordTok{install.packages}\NormalTok{(}\StringTok{"graphsim"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{#load required packages (once per R instance)}

\KeywordTok{library}\NormalTok{(}\StringTok{"igraph"}\NormalTok{)}
\CommentTok{#> }
\CommentTok{#> Attaching package: 'igraph'}
\CommentTok{#> The following objects are masked from 'package:stats':}
\CommentTok{#> }
\CommentTok{#>     decompose, spectrum}
\CommentTok{#> The following object is masked from 'package:base':}
\CommentTok{#> }
\CommentTok{#>     union}
\KeywordTok{library}\NormalTok{(}\StringTok{"graphsim"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(}\StringTok{"gplots"}\NormalTok{)}
\CommentTok{#> }
\CommentTok{#> Attaching package: 'gplots'}
\CommentTok{#> The following object is masked from 'package:stats':}
\CommentTok{#> }
\CommentTok{#>     lowess}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{#generate graph structure}
